---
title: "Assessing bicep curls"
author: "T Gocheco"
date: "January 25, 2016"
output: html_document
---

## Overview
This study takes in the Weight Lifting Exercise Dataset from http://groupware.les.inf.puc-rio.br/har and generates an analytics model that predicts whether a dumbbell lift has been done correctly or not given different measurements taken from some fitness measurement devices. Data has been downloaded from below links:  
- https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv  
- https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv  

In this study, the pml-training.csv data will be split into a training and a test set, and the training set will be used to develop a prediction model, and then tested on the test set. Finally once the results are satisfactory, the model will be used to predict the 20 observations of data from the pml-testing.csv  file, and properly classify whether the bicep curls were done correctly (classe A), or incorrectly as classified by classe B, C, D or E.  


## Data Loading.  
Loading training and testing data from files already downloaded. During loading, NA and blank fields are read as NA.  

```{r cache=TRUE}
pmltraining <- read.csv("./pml-training.csv", na.strings=c("NA", "#DIV/0!", ""));
pmltesting<- read.csv("./pml-testing.csv", na.strings=c("NA", "#DIV/0!", ""))
```

## Exploratory Data Analysis

Exploration done by doing a summary of the dataset. Around 100 variables have mostly NAs as values. See Appendix section, figure 1 for the summary.  

## Data processing and cleaning.  

Initial exploration shows 100 variables with mostly NA values. Following steps taken to identify the variables with NA, and removing them from the training and testing dataset.  
Also removing the first 7 columns which will not be considered as predictors in this study. These are timestamps and identifiers as to who was the subject, hence isn't really a predictor.  

```{r cache=TRUE}
library(caret)
b <-as.data.frame(summary(pmltraining))
NAlist <- b[grep('*NA*', b$Freq),]$Var2
NAlist<- sub("^\\s+", "", NAlist)

pmltrainingsubset <- pmltraining[,-which(names(pmltraining) %in% NAlist)]
pmltestingsubset <- pmltesting[,-which(names(pmltraining) %in% NAlist)]
pmltrainingsubset <- pmltrainingsubset[,-c(1:7)]
pmltestingsubset <- pmltestingsubset[,-c(1:7)]

```


## Prepare cross-validation data. 

Partition the original training data into 70% training and 30% test. 

```{r cache=TRUE}
set.seed(507)

inTrain<- createDataPartition(y=pmltrainingsubset$classe, p=0.7, list=FALSE)
training <- pmltrainingsubset[inTrain, ]; 
testing <- pmltrainingsubset[-inTrain,]
```


## Random forest Model creation.  

Chosen model is random forest. Though random forest already has some built-in cross-validation given the nature of the algorithm, this study will still do a further test on the test set once a model is created with the training set.  

Train the rainforest model with the training dataset with 53 variables, using default parameters of train.   

```{r cache=TRUE}
rffit <- train(classe ~ . , data=training, method="rf")
```


## Reviewing the result model.  


Viewing the resulting rain forest model:   

```{r}
rffit$finalModel
```

The out of sample error is estimated to be only __0.68%__ from the rainforest model. The confusion matrix returned by the best fit model seems to suggest an accuracy of __>98%__ across all 5 classes.  

Now let's view the variable importance plot. The variable importance plot below shows that rollbelt, pitchforearm, and yawbelt variables have the most influence as predictors in our model.   


```{r}
varImpPlot(rffit$finalModel, type=2, pch=19, col=1, cex=.5, main="")
```


## Cross Validation of model using the 30% testing dataset.  

We can now try the model on our test dataset as cross validation to check if the rainforest model performs as well with the testing dataset.  

```{r}
testprediction <- predict (rffit, testing)

confusionMatrix(testprediction, testing$classe)$table
```

Confusion matrix for test set vs model prediction shows good prediction across all 5 classes as well.  
Accuracy of the fit with the test set shows 99.08%, or an error of 0.92%.  

```{r}
confusionMatrix(testprediction, testing$classe)$overall
```

## Prediction of final test dataset  

Predicting the 20 test data of the pmltesting data using the rainforest model. 

```{r}
pmltestingpred <- predict(rffit, pmltestingsubset[,-53])
print(pmltestingpred)
```

Inputting the results into the final prediction quiz shows that all 20 exercises were successfully predicted by the model.  


## Appendix

### Figure 1: Summary of training dataset showing NA values. 

```{r}
summary(pmltraining)
```

### Credits.  

Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. Qualitative Activity Recognition of Weight Lifting Exercises. Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human '13) . Stuttgart, Germany: ACM SIGCHI, 2013.  

Read more: http://groupware.les.inf.puc-rio.br/har#ixzz3yGi7o6vv  
